Reproducibility of SNV-calling in multiple sequencing runs from single tumors

### 2015-10-01 PAPER/paper.tex
finished edits 10-11/13

Finally, we looked at the individual effects of the six of eight filters on the difference between replicates of each sample (only those mutations found in WGS \textit{or} WGA, \textit{but not} both), and on the similarity of WGS to WGA as a whole. We first looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA, where a higher ratio indicates that more mutations are removed from the difference than the overlap. We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}). To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}). We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order. (We did no look at 1the <10\% filter for this analysis, because it did not catch enough data to be meaningful.)

We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient (union/intersection, where the intersection is all those putative SNVs found in WGS \textit{or} WGA \textit{or} both) before and after filtering (Jaccard after-Jaccard before /Jaccard before). By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}). This is most likely because these two filters remove so much data: half or more of the overlap and as much or more of the difference. Removing so much of the data makes the intersection much, much smaller, therefor making the similarity between samples much larger, mostly independently of the size of the overlap.

%% this is figures 6
Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most of the overlap, was more likely to remove overlap for samples that showed a lot of overlap. We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend. Either filter was removing between 0 and 100\% of the overlap for some samples. For samples with more overlap, however, LOH was the primary filter removing overlap.

We found an additional trend that was initially unexpected: The LOH graph (Figure~\ref{fig:LOH-VAQ_all}) and the VAQ graph (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other. Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases. In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.

@@

After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates. We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository). The eight filters were enacted as follows:

@@

%% Figure 4
\begin{figure}
\centerline{
\includegraphics[width=6in]{Figure4.pdf} }
\caption{\textbf{Different filters catch greater or smaller numbers of putative SNVs, and lower or higher percentages of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample. The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap. The <10\% filter removed very few putative SNVs and almost none of the overlap. The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data. }
\label{fig:boxplot_filtered}
\end{figure}

%%Figure 5
\begin{figure}
\centerline{
\includegraphics[width=5in]{Figure5.pdf} }
\caption{\textbf{A title.} a caption.}
\label{fig:diffset}
\end{figure}

>>>>

We next looked at the individual effects of the six of eight filters on the difference between replicates of each sample, those mutations found only in WGS \textit{or} WGA, \textit{but not} both. We looked at the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$), where a higher ratio indicates that more mutations are removed from the difference than the overlap. We first plotted this ratio for each sample for each filter (Figure~\ref{fig:diffset}, \textit{top left}). To better compare the filters to each other, we looked at the distribution of ratios across samples for each filter (Figure~\ref{fig:diffset}, \textit{top right}). We found that the LOH and VAQ filters scored worst, followed by the 10bp-SNV, 10bp-INDEL, and dbSNP filters, in that order. (We did no look at the <10\% filter for this analysis, because it did not catch enough data to be meaningful.) Thus, those filters that remove the fewest putative SNVs in the overlap, also remove the highest number of SNVs in the difference relative to the overlap.

We next looked at the similarity of WGS to WGA as a whole as measured by the difference, normalized to 100, in the Jaccard similarity coefficient ($WGS \cap WGA / WGS \cup WGA$) before and after filtering (Jaccard after-Jaccard before /Jaccard before). By this measure, and in contrast to the previous two measures, we found that LOH and VAQ performed the best by an order of magnitude, followed by 10bp-SNV, 10bp-INDEL, and dbSP, in that order (Figure~\ref{fig:diffset}, \textit{below}). We see this effect because these two filters remove so much data: half or more of the overlap and as much or more of the difference. Removing so much of the data makes the intersection much, much smaller, and thus makes the Jaccard coefficient between samples much larger, mostly independently of the size of the overlap.

%% this is figure 6
Finally, we asked whether either the VAQ or the LOH filter, responsible between them for removing most to all of the overlap, was more likely to remove overlap for samples that showed a lot of overlap. We plotted the percent of the overlap filtered out by the LOH filter (Figure~\ref{fig:LOH-VAQ_all}) and the percent of the overlap filtered out by the VAQ filter (Figure~\ref{fig:LOH-VAQ_all}) against the number of putative mutations in the overlap, and we found that for samples with overlap of $\lesssim 100$ there was no strong trend. Either filter was removing between 0 and 100\% of the overlap for some samples. For samples with more overlap, however, LOH was the primary filter removing overlap.

We found an additional trend that was initially unexpected: The LOH and VAQ graphs (Figure~\ref{fig:LOH-VAQ_all}) looked like exact inverses of each other. Further inspection showed that the sum of the percent overlap removed by LOH and by VAQ was nearly, but not exactly, 100\% in all cases. In hindsight, this result was somewhat expected, since (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time.

@@

After generating a VCF file with all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates. We then filtered the lists of putative SNVs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository). Each of the eight filters was run independently on the whole data set. The eight filters were enacted as follows:

@@

%% Figure 4
\begin{figure}
\centerline{
\includegraphics[width=6in]{Figure4.pdf} }
\caption{\textbf{Independent performance of the individual filters of the overlap between replicates.} The $x$-axis gives the name of each filter (detail on Table 1), and the $y$-axis gives \textit{(above)} the number of mutations removed by a given filter in a given sample on a log scale, and \textit{(below)} the percentage of the WGS--WGA overlap removed by the filter, per sample. The LOH and VAQ filters removed a large number of putative SNVs and portion of the overlap. The <10\% filter removed very few putative SNVs and almost none of the overlap. The 10bp-SNV, 10bp-INDEL, and dbSNP filters removed an intermediate number of putative SNVs (~100), but only a very small portion of the overlap, making them the best performers on the overlap data. }
\label{fig:boxplot_filtered}
\end{figure}

%%Figure 5
\begin{figure}
\centerline{
\includegraphics[width=5in]{Figure5.pdf} }
\caption{\textbf{Independent performance of individual filters of the difference between replicates.} \textit{(top left)} A scatter plot of the ratio, per sample per filter, of putative SNVs removed from the difference versus the overlap of WGS and WGA ($WGS \triangle WGA / WGS \cap WGA$). \textit{(top right)} Boxplot of the same data, divided by filter on the $x$-axis. \textit{(below)} Plot of the percentage difference in the Jaccard(WGS, WGA) before and after the action of each filter (each filter was run on the whole data set independently).}
\label{fig:diffset}
\end{figure}
